Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-01-25 xrb2/paper.tex
De-blue text

Burning in the {\tt F500} and {\tt F1000} runs is concentrated in a dense region with circular 
motion. In Figure~\ref{fig:urho}, which compares the horizontal $r$-velocity $u$, density $\rho$ and the 
nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the {\tt F250}, {\tt F500}, and {\tt F1000} 
runs, most of the burning for each of the simulations occurs in a high density region $\rho > 3 
\times 10^5 {\rm g\,cm^{-3}}$. The fluid in this dense, high energy generation region undergoes 
vortical motion, shown in the Figure~\ref{fig:uv} phase plots comparing $u$, the vertical $z$-velocity $v$ and $\dot{e}_{\rm 
nuc}$. This most likely corresponds to the leading edge of the flame where fresh fuel is being entrained. \MarginPar{not sure if we should keep this sentence about entrainment...}
This feature is not developed in the $250~\mathrm{Hz}$ flame in Figure~\ref{fig:uv} {\color{blue} (left panel)}; it could potentially develop 
at later times (past the point at which we terminated our simulation), or the flame could just 
fizzle out and fail to ignite entirely.

@@

Figure~\ref{fig:compare_500Hz_abar} shows $\bar{A}$ for three $500~\mathrm{Hz}$ simulations with different initial temperature structures ($T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$) at $t = 70~\mathrm{ms}$. {\color{blue}We do not plot $\tt{F500\_4E8}$ here because it fails to form a clear burning front. $\tt{F500\_3E8}$ (Figure~\ref{fig:compare_500Hz_abar}, bottom panel) --- the hottest run to form a clear burning front ---} has a faster propagating flame (this will be discussed further in Section \ref{ssec:temp_prop}). {\color{blue} It also reaches slightly higher $\bar{A}$ values than the two cooler runs. The Figure~\ref{fig:abar_hot} $\bar{A}$-$T$ phase plots of $\tt{F500}$ (left) and $\tt{F500\_3E8}$ (middle) also reflect these $\bar{A}$ features, with $\tt{F500\_3E8}$ reaching slightly higher $\bar{A}$ values. $\tt{F500\_3E8}$ also reaches higher $\enucdot$ values at the low end of the temperature range ($< 0.5 \times 10^9 \mathrm{K}$). There appear to be more causally connected regions across a range of $\bar{A}$ at low temperatures for $\tt{F500\_3E8}$ than for $\tt{F500}$, suggesting that the higher $\enucdot$ for $\tt{F500\_3E8}$ generates burning in certain reaction pathways that are not present in the cooler $\tt{F500}$ run. Note that, although $\tt{F500\_3E8}$ is the hottest run with a modified initial temperature configuration to form a distinct flame front, the highest rotation $\tt{F1000}$ flame actually reaches higher temperatures (see Figure~\ref{fig:maxima}, left panel) as well as higher $\bar{A}$ (see Figure~\ref{fig:time_series_1000}).}

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ 
is so hot that the organized flame structure is lost. The fuel layer undergoes stable burning and 
essentially simmers in place, rather than forming a flame. This process has been observed in 
neutron stars with accretion rates that exceed the Eddington limit 
\citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}; in this model we have essentially 
recreated the conditions found in such stars with high accretion rates. This model burns so 
strongly that it is only run out to $40~\mathrm{ms}$. After an initial period when the burning 
moves across the domain, residual burning continues and eventually ignites the entire fuel layer at 
late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three 
snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, 
$\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ 
across the domain and at heights up to $\sim 0.5\times 10^4~\mathrm{cm}$. There is still 
significant burning occurring even higher, with $\enucdot$ reaching $\sim 
10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 
10^4~\mathrm{cm}$. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum 
$\enucdot$ values on the order of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ 
{\color{blue} (see Figure~\ref{fig:maxima}, right panel)}, even at the latest timesteps ($\sim 
70~\mathrm{ms}$, when the flame is most developed). {\color{blue} Significant $\enucdot$ values for 
all other runs are confined to the flame front, rather than spanning the entire domain}.

@@

 {\color{blue} Though $\tt{F500\_4E8}$ does not form a distinct burning front, it does achieve greatly enhanced burning, as shown in the Figure~\ref{fig:abar_hot} $\bar{A}$-$T$ phase plot (right panel). Not only is $\tt{F500\_4E8}$ significantly hotter overall compared to all other runs}, but there is a large causally connected region across a wide range of $\bar{A}$. This indicates that the hotter temperature has facilitated significant burning in reaction pathways that were not favored at the lower temperatures. The reaction pathways are also very disrupted for $\tt{F500\_4E8}$ compared to the cooler runs, suggesting that the hotter temperature leads to more vigorous mixing. {\color{blue}Indeed, this appears to be the case looking at Figure~\ref{fig:4e8_stacked_abar} and the right panel of Figure~\ref{fig:uv_hot} (discussed further in section~\ref{ssec:temp_prop}). The disrupted reaction pathways resemble those found in Figure \ref{fig:abar} for the highest rotation $\tt{F1000}$ run, though they are even more dramatically disrupted for $\tt{F500\_4E8}$. $\tt{F500\_4E8}$ is also noticeably hotter than $\tt{F1000}$, even though it is run for significantly less time ($t = 40~\mathrm{ms}$ vs $t = 100~\mathrm{ms}$). Although the $\tt{F500\_4E8}$ run is clearly a special case in that it develops steady burning across the domain rather than a propagating flame, enhanced burning in this hottest run aligns with results from the other simulations with differing initial temperature structures. As $T_{\mathrm{star}}$ is increased, the overall behavior and propagation of the flame is significantly altered, implying that burning is very temperature-sensitive. We explore flame propagation further in Section~\ref{ssec:temp_prop}.}

@@

The $u$-$\rho$ phase plot in Figure~\ref{fig:urho_hot} shows that the horizontal velocity for the $T_{\rm star} = 4\times10^8$~K {\color{blue} ($\tt{F500\_4e8}$; right panel)} is significantly greater in magnitude than for the cooler {\color{blue}$\tt{F500\_3e8}$ (middle panel) and $\tt{F500}$ (left panel)} runs. The $u$-$v$ plot in Figure~\ref{fig:uv_hot} further demonstrates this, as again we see that the magnitude of the {\color{blue}$\tt{F500\_4e8}$ flame's vertical $z$-velocity $v$} is also significantly larger. This motion therefore most likely corresponds to the vortical motion rather than to the propagation of the flame itself, something which is further supported by the fact that the velocity magnitude is two to three orders of magnitude greater than the flame speeds listed in Table~\ref{table:flame_speeds}. Also of interest in these plots is that the coolest run does not seem to have developed the characteristic vortex structure at the burning front (i.e. where $\dot{e}_{\rm nuc}$ is greatest) that can be clearly seen for the hotter runs at this time ($40~\mathrm{ms}$). As can be seen in Figure~\ref{fig:uv} (which was plotted at $100~\mathrm{ms}$), this does develop more at later times. Similar to what we saw when comparing the runs with different rotation rates, it would therefore appear that it takes longer for the flame to develop when $T_{\rm star}$ is cooler.

\begin{figure}[t]
    \centering
    \plotone{urho_plot_hotter}
    \caption{\label{fig:urho_hot}Phase plot of the horizontal $r$-velocity $u$, density $\rho$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$ {\color{blue} (the latest time that $\tt{F500\_4E8}$ is run out to). The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.}
\end{figure}

\begin{figure}[t]
    \centering
    \plotone{uv_plot_hotter}
    \caption{\label{fig:uv_hot}Phase plot of the horizontal $r$-velocity $u$, vertical $z$-velocity $v$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$. {\color{blue} The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.}
\end{figure}

\begin{figure}[t]
    \centering
    \plotone{abar_plot_hotter}
    \caption{\label{fig:abar_hot}Phase plot of the mean molecular weight $\bar{A}$, temperature $T$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$. {\color{blue} The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.}
\end{figure}

>>>>

Burning in the {\tt F500} and {\tt F1000} runs is concentrated in a dense region with circular 
motion. In Figure~\ref{fig:urho}, which compares the horizontal $r$-velocity $u$, density $\rho$ and the 
nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the {\tt F250}, {\tt F500}, and {\tt F1000} 
runs, most of the burning for each of the simulations occurs in a high density region $\rho > 3 
\times 10^5 {\rm g\,cm^{-3}}$. The fluid in this dense, high energy generation region undergoes 
vortical motion, shown in the Figure~\ref{fig:uv} phase plots comparing $u$, the vertical $z$-velocity $v$ and $\dot{e}_{\rm 
nuc}$. This most likely corresponds to the leading edge of the flame where fresh fuel is being entrained. \MarginPar{not sure if we should keep this sentence about entrainment...}
This feature is not developed in the $250~\mathrm{Hz}$ flame in Figure~\ref{fig:uv} (left panel); it could potentially develop 
at later times (past the point at which we terminated our simulation), or the flame could just 
fizzle out and fail to ignite entirely.

@@

Figure~\ref{fig:compare_500Hz_abar} shows $\bar{A}$ for three $500~\mathrm{Hz}$ simulations with different initial temperature structures ($T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$) at $t = 70~\mathrm{ms}$. We do not plot $\tt{F500\_4E8}$ here because it fails to form a clear burning front. $\tt{F500\_3E8}$ (Figure~\ref{fig:compare_500Hz_abar}, bottom panel) --- the hottest run to form a clear burning front --- has a faster propagating flame (this will be discussed further in Section \ref{ssec:temp_prop}). It also reaches slightly higher $\bar{A}$ values than the two cooler runs. The Figure~\ref{fig:abar_hot} $\bar{A}$-$T$ phase plots of $\tt{F500}$ (left) and $\tt{F500\_3E8}$ (middle) also reflect these $\bar{A}$ features, with $\tt{F500\_3E8}$ reaching slightly higher $\bar{A}$ values. $\tt{F500\_3E8}$ also reaches higher $\enucdot$ values at the low end of the temperature range ($< 0.5 \times 10^9 \mathrm{K}$). There appear to be more causally connected regions across a range of $\bar{A}$ at low temperatures for $\tt{F500\_3E8}$ than for $\tt{F500}$, suggesting that the higher $\enucdot$ for $\tt{F500\_3E8}$ generates burning in certain reaction pathways that are not present in the cooler $\tt{F500}$ run. Note that, although $\tt{F500\_3E8}$ is the hottest run with a modified initial temperature configuration to form a distinct flame front, the highest rotation $\tt{F1000}$ flame actually reaches higher temperatures (see Figure~\ref{fig:maxima}, left panel) as well as higher $\bar{A}$ (see Figure~\ref{fig:time_series_1000}).

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ 
is so hot that the organized flame structure is lost. The fuel layer undergoes stable burning and 
essentially simmers in place, rather than forming a flame. This process has been observed in 
neutron stars with accretion rates that exceed the Eddington limit 
\citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}; in this model we have essentially 
recreated the conditions found in such stars with high accretion rates. This model burns so 
strongly that it is only run out to $40~\mathrm{ms}$. After an initial period when the burning 
moves across the domain, residual burning continues and eventually ignites the entire fuel layer at 
late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three 
snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, 
$\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ 
across the domain and at heights up to $\sim 0.5\times 10^4~\mathrm{cm}$. There is still 
significant burning occurring even higher, with $\enucdot$ reaching $\sim 
10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 
10^4~\mathrm{cm}$. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum 
$\enucdot$ values on the order of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ 
(see Figure~\ref{fig:maxima}, right panel), even at the latest timesteps ($\sim 
70~\mathrm{ms}$, when the flame is most developed). Significant $\enucdot$ values for 
all other runs are confined to the flame front, rather than spanning the entire domain.

@@

 Though $\tt{F500\_4E8}$ does not form a distinct burning front, it does achieve greatly enhanced burning, as shown in the Figure~\ref{fig:abar_hot} $\bar{A}$-$T$ phase plot (right panel). Not only is $\tt{F500\_4E8}$ significantly hotter overall compared to all other runs, but there is a large causally connected region across a wide range of $\bar{A}$. This indicates that the hotter temperature has facilitated significant burning in reaction pathways that were not favored at the lower temperatures. The reaction pathways are also very disrupted for $\tt{F500\_4E8}$ compared to the cooler runs, suggesting that the hotter temperature leads to more vigorous mixing. Indeed, this appears to be the case looking at Figure~\ref{fig:4e8_stacked_abar} and the right panel of Figure~\ref{fig:uv_hot} (discussed further in section~\ref{ssec:temp_prop}). The disrupted reaction pathways resemble those found in Figure \ref{fig:abar} for the highest rotation $\tt{F1000}$ run, though they are even more dramatically disrupted for $\tt{F500\_4E8}$. $\tt{F500\_4E8}$ is also noticeably hotter than $\tt{F1000}$, even though it is run for significantly less time ($t = 40~\mathrm{ms}$ vs $t = 100~\mathrm{ms}$). Although the $\tt{F500\_4E8}$ run is clearly a special case in that it develops steady burning across the domain rather than a propagating flame, enhanced burning in this hottest run aligns with results from the other simulations with differing initial temperature structures. As $T_{\mathrm{star}}$ is increased, the overall behavior and propagation of the flame is significantly altered, implying that burning is very temperature-sensitive. We explore flame propagation further in Section~\ref{ssec:temp_prop}.

@@

The $u$-$\rho$ phase plot in Figure~\ref{fig:urho_hot} shows that the horizontal velocity for the $T_{\rm star} = 4\times10^8$~K ($\tt{F500\_4e8}$; right panel) is significantly greater in magnitude than for the cooler $\tt{F500\_3e8}$ (middle panel) and $\tt{F500}$ (left panel) runs. The $u$-$v$ plot in Figure~\ref{fig:uv_hot} further demonstrates this, as again we see that the magnitude of the $\tt{F500\_4e8}$ flame's vertical $z$-velocity $v$ is also significantly larger. This motion therefore most likely corresponds to the vortical motion rather than to the propagation of the flame itself, something which is further supported by the fact that the velocity magnitude is two to three orders of magnitude greater than the flame speeds listed in Table~\ref{table:flame_speeds}. Also of interest in these plots is that the coolest run does not seem to have developed the characteristic vortex structure at the burning front (i.e. where $\dot{e}_{\rm nuc}$ is greatest) that can be clearly seen for the hotter runs at this time ($40~\mathrm{ms}$). As can be seen in Figure~\ref{fig:uv} (which was plotted at $100~\mathrm{ms}$), this does develop more at later times. Similar to what we saw when comparing the runs with different rotation rates, it would therefore appear that it takes longer for the flame to develop when $T_{\rm star}$ is cooler.

\begin{figure}[t]
    \centering
    \plotone{urho_plot_hotter}
    \caption{\label{fig:urho_hot}Phase plot of the horizontal $r$-velocity $u$, density $\rho$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$ (the latest time that $\tt{F500\_4E8}$ is run out to). The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.
\end{figure}

\begin{figure}[t]
    \centering
    \plotone{uv_plot_hotter}
    \caption{\label{fig:uv_hot}Phase plot of the horizontal $r$-velocity $u$, vertical $z$-velocity $v$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$. The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.
\end{figure}

\begin{figure}[t]
    \centering
    \plotone{abar_plot_hotter}
    \caption{\label{fig:abar_hot}Phase plot of the mean molecular weight $\bar{A}$, temperature $T$ and the nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the $500~\mathrm{Hz}$ runs with $T_{\rm star}$ of $10^8~\mathrm{K}$, $3\times10^8~\mathrm{K}$ and $4\times10^8~\mathrm{K}$ ($\tt{F500}$, $\tt{F500\_3e8}$, and $\tt{F500\_4e8}$) at $t = 40~\mathrm{ms}$. The run with $T_{\rm star} = 2\times10^8~\mathrm{K}$ ($\tt{F500\_2E8}$) closely resembles the cooler $\tt{F500}$ run, and is not shown here}.
\end{figure}
